Adversarially Robust Medical Classification via Attentive Convolutional Neural Networks

### 2022-07-25 paper/PaperForReview.tex
Working  on discussion

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{lipsum}
\usepackage{tikz}
\usepackage{relsize}

@@

    \subsection{Network Architecture}
      Two models were trained for each dataset, one with a soft attention block and one without. The architecture of each model was based on the TensorFlow \cite{TensorFlow} implementation of ResNet-50 \cite{ResNetImplementation}. ResNet-50 was instantiated with the top block included, initial weights based on ImageNet \cite{ImageNet} pretraining, and class count set to 1000. The final three layers of the network were removed. In the models without attention, these final layers were replaced with a 2D global average pooling layer followed by a fully-connected layer with softmax activation. In the models with attention, the last three layers were replaced with a soft attention block implemented according to the specifications of Datta et al. (2021). This block produces a $7 \times 7$ feature map which is $2 \times 2$ max pooled and concatenated with a $2 \times 2$ max pooled version of the input to the attention block. This concatenated output is put through ReLU activation, 50\% dropout, and global average pooling before being handed off to the fully connected prediction head with softmax activation \cite{AttentionSkinCancerClassification}. In each model, the width of the prediction head was equal to the number of classes in the dataset (seven for dermatoscopy and two for fundoscopy and chest x-ray).

@@

  \subsection{Evaluation of Robustness}

@@

>>>>

% Include other packages here, before hyperref.
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{lipsum}
\usepackage{tikz}
\usepackage{relsize}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\pgfplotsset{compat=1.7}

@@

    \subsection{Network Architecture}
      Four models were trained for each dataset, a ResNet-50 with and without a soft attention block and an InceptionResNetV2 with and without soft attention. The architecture of each model was based on the TensorFlow \cite{TensorFlow} implementations of ResNet-50 \cite{ResNetImplementation} and InceptionResNetV2 \cite{InceptionImplementation}.

      ResNet-50 was instantiated with the top block included, initial weights based on ImageNet \cite{ImageNet} pretraining, and class count set to 1000. The final three layers of the network were removed. In the models without attention, these final layers were replaced with a 2D global average pooling layer followed by a fully-connected layer with softmax activation. In the models with attention, the last three layers were replaced with a soft attention block implemented according to the specifications of Datta et al. (2021). This block produces a $7 \times 7$ feature map which is $2 \times 2$ max pooled and concatenated with a $2 \times 2$ max pooled version of the input to the attention block. This concatenated output is put through ReLU activation, 50\% dropout, and global average pooling before being handed off to the fully connected prediction head with softmax activation \cite{AttentionSkinCancerClassification}.

      InceptionResNetV2 \cite{InceptionResNet} was instantiated with the top block included, initial weights based on ImageNet, and classifier activation as softmax. In the models without attention, the final 28 layers were replaced with a ReLU activation and 50\% dropout whose output was flattened and sent to the fully connected prediction head with softmax activation. In the models with attention, these layers were replaced by the soft attention block described above. This output is $2 \times 2$ max pooled and concatenated with a max pooled version of the input to the attention block before being sent through ReLU activation and the softmax prediction head.

      In each model, the width of the prediction head was equal to the number of classes in the dataset (seven for dermatoscopy and two for fundoscopy and chest x-ray).

@@

    \subsection{Evaluation and Analysis of Robustness}
      After training, the models' clean- and robust-accuracy was evaluated using the FoolBox \cite{FoolBoxPaper}\cite{FoolBoxLibrary} library's implementation of $l_\infty$ projected gradient descent attacks. Each set of models was tested at increasing epsilons (perturbation radii). Attacks of these perturbation radii were created for each image in the test sets. Unweighted accuracy was calculated for each perturbation radius.

      Small samples of the resultant adversarial images were saved and used to analyze model behavior using gradient weighted class activation mapping (Grad-CAM) \cite{Grad-CAM} localization maps. These maps were generated using the Grad-CAM implementation of Datta et al. (2021) \cite{AttentionSkinCancerClassification} and provide attention-map-esque representations of spatial importance for neural networks regardless of whether they utilize attention mechanisms or not. For each model, these maps were generated at each epsilon for 16 sample images with and without perturbations applied. Maps for a given starting image were compared to their perturbed counterpart as well as against other models and perturbation radii.

  \section{Results}
    \subsection{ResNet-50 Models}
      For the task of skin lesion classification, the models with attention are clearly superior, in terms of robustness (Fig. \ref{DermResNet50Robustness}). Although the model without attention has slightly higher clean-accuracy (0.905 vs. 0.88), even the slightest perturbation ($\epsilon=0.00125$) is able to reduce its accuracy by 24\% and put the model with attention in the lead. By the time $\epsilon=0.005$, the model is worse than random, and by the time $\epsilon=0.01$, the model is rendered completely useless. It is worth noting that perturbation of this size (Fig. \ref{example_derm_perturbation}) is far from being human perceptible. Meanwhile, at this perturbation radius, the model with attention remains better than random selection.

      \begin{figure}[h!]
        \begin{tikzpicture}
          \begin{axis}[xtick={0, 0.00125, 0.0025, 0.005, 0.01, 0.02, 0.04, 0.08, 0.16, 0.32}, x tick label style={rotate=45, log ticks with fixed point},xmode=log, log basis x=2, xlabel=Perturbation Radius ($\epsilon$), ylabel=Accuracy, width=\linewidth, height=7cm,legend style={at={(0.95,0.95)},anchor=north east}]

          \addplot[color=red,mark=x] coordinates {
            (0.00125, 0.687)
            (0.0025, 0.216)
            (0.005, 0.012)
            (0.01, 0)
          };

          \addplot[color=blue,mark=*] coordinates {
            (0.00125, 0.795)
            (0.0025, 0.685)
            (0.005, 0.337)
            (0.01, 0.022)
          };

          \addplot[color=red, domain=0.00125:0.01, dashed]{0.905};
          \addplot[color=blue, domain=0.00125:0.01, dashed]{0.880};

          \legend{Without Attention,With Attention}
          \end{axis}
          \end{tikzpicture}
        \caption{Clean- and robust-accuracy of ResNet-50 models for skin lesion classification.  Dashed lines represent clean-accuracy.}
        \label{DermResNet50Robustness}
      \end{figure}

      The models for diabetic retinopathy detection tell a slightly different story (Fig. \ref{DRResNet50Robustness}). These models were, overall, much more robust, requiring a perturbation of at least $\epsilon=0.01$ to reduce accuracy by 3\%. As in the skin lesion classification models, the model without attention has a slightly higher clean accuracy (0.832 vs. 0.818). It retains this very small lead until $\epsilon=0.02$, at which point, the accuracy of both models begins falling, but the model with attention does so a bit less dramatically. At the maximum perturbation radius tested ($\epsilon=0.32$) (Fig. \ref{example_dr_perturbation}), the accuracy of the model with attention is over 20 times higher than that of the model without.

      \begin{figure}[h!]
        \begin{tikzpicture}
          \begin{axis}[xtick={0, 0.00125, 0.0025, 0.005, 0.01, 0.02, 0.04, 0.08, 0.16, 0.32}, x tick label style={rotate=45, log ticks with fixed point},xmode=log, log basis x=2, xlabel=Perturbation Radius ($\epsilon$), ylabel=Accuracy, width=\linewidth, height=7cm,legend style={at={(0.05,0.05)},anchor=south west}]

          \addplot[color=red,mark=x] coordinates {
            (0, 0.832)
            (0.00125, 0.828)
            (0.0025, 0.825)
            (0.005, 0.819)
            (0.01, 0.806)
            (0.02, 0.778)
            (0.04, 0.705)
            (0.08, 0.506)
            (0.16, 0.175)
            (0.32, 0.014)
          };

          \addplot[color=blue,mark=*] coordinates {
            (0, 0.818)
            (0.00125, 0.816)
            (0.0025, 0.815)
            (0.005, 0.813)
            (0.01, 0.808)
            (0.02, 0.8)
            (0.04, 0.780)
            (0.08, 0.733)
            (0.16, 0.599)
            (0.32, 0.287)
          };

          \addplot[color=red, domain=0.00125:0.32, dashed]{0.832};
          \addplot[color=blue, domain=0.00125:0.32, dashed]{0.818};

          \legend{Without Attention,With Attention}
          \end{axis}
          \end{tikzpicture}
        \caption{Clean- and robust-accuracy of ResNet-50 models for diabetic retinopathy detection. Dashed lines represent clean-accuracy.}
        \label{DRResNet50Robustness}
      \end{figure}

      The pneumothorax detection models, collectively, were even more robust than the diabetic retinopathy detection models (Fig. \ref{PneumoResNet50Robustness}). After perturbations of $\epsilon=0.32$ (Fig. \ref{example_pneumo_perturbation}), the accuracy of both models met at 0.805. Prior to this, the accuracy of the model without attention hovered 1-2\% above that of the model with attention. However, based on the trajectory of the models' accuracy with respect to $\epsilon$, perturbation radii higher than 0.32 would likely result in the model with attention leapfrogging the model without.

      \begin{figure}[h!]
        \begin{subfigure}{\linewidth}
          \centering
          $
          \begin{array}{l}
          \includegraphics[width=0.26\linewidth]{graphics/derm_unperturbed.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{+}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/derm_e=0.01_difference.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{=}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/derm_e=0.01.jpg}
          \end{array}
          $
          \caption{An example of $\epsilon=0.01$ perturbation on the dermatoscopy dataset. 0.01 is the maximum perturbation radius used for this dataset.\\}
          \label{example_derm_perturbation}
        \end{subfigure}

        \begin{subfigure}{\linewidth}
          \centering
          $
          \begin{array}{l}
          \includegraphics[width=0.26\linewidth]{graphics/dr_unperturbed.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{+}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/dr_e=0.32_difference.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{=}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/dr_e=0.32.jpg}
          \end{array}
          $
          \caption{An example of $\epsilon=0.32$ perturbation on the fundoscopy dataset. 0.32 is the maximum perturbation radius used for this dataset.\\}
          \label{example_dr_perturbation}
        \end{subfigure}
        \begin{subfigure}{\linewidth}
          \centering
          $
          \begin{array}{l}
          \includegraphics[width=0.26\linewidth]{graphics/pneumo_unperturbed.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{+}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/pneumo_e=0.32_difference.jpg}
          \end{array}
          \mathlarger{\mathlarger{\mathlarger{\mathlarger{=}}}}
          \begin{array}{l}
            \includegraphics[width=0.26\linewidth]{graphics/pneumo_e=0.32.jpg}
          \end{array}
          $
          \caption{An example of $\epsilon=0.32$ perturbation on the chest x-ray dataset. 0.32 is the maximum perturbation radius used for this dataset.\\}
          \label{example_pneumo_perturbation}
        \end{subfigure}%
        \caption{Perturbation examples}
      \end{figure}
      \begin{figure}[h!]
        \begin{tikzpicture}
          \begin{axis}[xtick={0, 0.00125, 0.0025, 0.005, 0.01, 0.02, 0.04, 0.08, 0.16, 0.32}, x tick label style={rotate=45, log ticks with fixed point},xmode=log, log basis x=2, xlabel=Perturbation Radius ($\epsilon$), ylabel=Accuracy, width=\linewidth, height=7cm,legend style={at={(0.05,0.05)},anchor=south west}]

          \addplot[color=red,mark=x] coordinates {
            (0.00125, 0.897)
            (0.0025, 0.897)
            (0.005, 0.896)
            (0.01, 0.896)
            (0.02, 0.894)
            (0.04, 0.889)
            (0.08, 0.881)
            (0.16, 0.864)
            (0.32, 0.805)
          };

          \addplot[color=blue,mark=*] coordinates {
            (0.00125, 0.883)
            (0.0025, 0.883)
            (0.005, 0.882)
            (0.01, 0.882)
            (0.02, 0.880)
            (0.04, 0.877)
            (0.08, 0.870)
            (0.16, 0.853)
            (0.32, 0.805)
          };

          \addplot[color=red, domain=0.00125:0.32, dashed]{0.897};
          \addplot[color=blue, domain=0.00125:0.32, dashed]{0.883};

          \legend{Without Attention,With Attention}
          \end{axis}
          \end{tikzpicture}
        \caption{Clean- and robust-accuracy of ResNet-50 models for pneumothorax detection. Dashed lines represent clean-accuracy.}
        \label{PneumoResNet50Robustness}
      \end{figure}

    \subsection{InceptionResNetV2 Models}
  \section{Discussion}
    \subsection{Analysis of Attention Maps}
    \subsection{Effects of Attention} % What happened? Did it work?
    \subsection{Conclusion} % Result, contribution, future work

@@

\pagebreak
\appendix
\section{Additional Figures}
